---
title: "Functions"
---

> **Sources for this section**
>
> [R Function Writing 101 \| R bloggers](https://www.r-bloggers.com/2024/01/r-function-writing-101a-journey-through-syntax-best-practices-and-more/)

# What is a function?

A **function** is a set of instructions packaged into a resusable block of code. We already use functions created by package developers in our everyday coding (i.e `read_excel()`, `clean_names()`, and `mean()`. This section will demonstrate how to create our own functions!

**Rule of Thumb**: If you are repeating the same or very similar code more than 2 times, it is more efficient to create a function for it.

**Structure of a function**

``` r
my_function <- function(arg1, arg2, ...) {      
  # function body      # perform instructions using arg1 and arg2      return(result)  }
```

1.  **Function Definition** `my_function`: The name you assign your function.
2.  **Keyword** `function()`: Tells R you are creating a new function.
3.  **Parameter** `(arg1, arg2, ...)`: Arguments (also called inputs) passed to the function. These can include:
    1.  Data the function will use (such as a dataset, column, number, or text).
    2.  Setting that change how the function behaves.
4.  **Return Statement** `return(result)`: The final result or output produced by the function.

**Example (level 1)**

```{r}
#create function  
square <- function(x) {
  result <- x^2      
  return(result)    
  }      
squared_value <- square(4) # usage of function  print(squared_value) # view result
```

1.  **Function Definition**
    -   `square` is the name assigned to the function.
2.  **Keyword**
    -   `function()` is telling R that you are creating a new function.
3.  **Parameter**
    -   `x` is the input value the function will perform calculations on. In this example, it represents the number you want to square.
4.  **Function Body**
    -   Any instructions that are enclosed in curly braces `{ }`.
    -   `result <- x^2` calculates the square of x.
5.  **Return Statement** - `return(result)` specifies the final output in the function will return.
6.  **Usage**
    -   `squared_value` is the assigned name for a new output (in this case, a variable).
    -   `square(4)` is applying the function on the value `4`.
        -   `4^2`
    -   `print(squared_value)` prints the result to the console `16`.

**Summary**: The function takes a single argument, squares it, and returns the result.

```{r}
squared_value <- square(18) # usage of function    
print(squared_value) # view result
```

**Example (level 2)**

Let's say, we want a function that calculates an exponent of a value but with flexibility - a single function that can square, cube, or a value to any power.

```{r}
power_function <- function(x, exponent = 2) {
  result <- x^exponent     
  return(result) 
  }    
power_deafult_value <- power_function(4)    
print(power_deafult_value) 
```

```{r}
power_option_value <- power_function(4, exponent = 3)   
print(power_option_value) 
```

In this example, we added a new **parameter (or argument)**.

1.  **Function Definition**
    -   `power_funtion` is the name assigned to the function.
2.  **Keyword**
    -   `function()` is telling R that you are creating a new function.
3.  **Parameter**
    -   `x` is the input value the function will perform calculations on. In this example, it represents the number you want to raise to a power.
    -   `exponent = 2` is the argument that controls how the function operates. The default value is `2`, meaning the function will square the input value unless another exponent is specified.
4.  **Function Body**
    -   Any instructions that are enclosed in curly braces `{ }`.
    -   `result <- x^exponent` calculates the value of `x` raised to the selected exponent.
5.  **Return Statement** - `return(result)` specifies the final output in the function will return.
6.  **Usage \# 1**
    -   `power_default_value` is the assigned name for a new output (in this case, a variable).
    -   `power_function(4)` is applying the function to the value `4`. Since no exponent was specified, the function uses the default value of `2`.
        -   `4^2`
    -   `print(power_default_value)` prints the result to the console `16`.
7.  **Usage \# 2**
    -   `power_option_value` is the assigned name for a new output (in this case, a variable).
    -   `power_function(4, exponent = 3)` is applying the function to the value `4` , while specifying an exponent of `3`.
        -   `4^3`
    -   `print(power_option_value)` prints the result to the console `64`.

There are other ways to define **arguments** but I want to focus more on the practical application rather than going to deeply into the computer science concepts.

# Practical functions

In this section, I've compiled some custom functions that may be useful to adapt or repurpose for your own projects. In each scenario, i've included a repurposing rating to indicate how worthwhile it is for adaption.

## Scenario 1 (⭐⭐/5)

It's that time of year where you perform similar analysis annually. You have separate datasets for *E. coli*, *Salmonella*, and *Campylobacter* and you want to select similar columns from each dataset - but with minor difference, such as an antimicrobial being used for one organism but not another. Rather than repeating most of your code three separate times, you can create a reusable function and apply it across datasets like this:

``` r
# create function that specifies relevant columns   
select_amr_columns <- function(df, organism) {
if (organism == "ecoli") { 
  df |>             
    select(  
      isolate_id,
      region,  
      year,   
      month, 
      qtr,  
      sttype,
      final_id_genus, 
      final_id_species, 
      cmic_amc, 
      cmic_amp,  
      cmic_azm, 
      cmic_fox, 
      cmic_cro, 
      cmic_chl, 
      cmic_cip, 
      cmic_gen, 
      cmic_nal,  
      cmic_sss, 
      cmic_tet, 
      cmic_sxt,      
      cmic_tio,  
      cmic_str  
      )        
  } else if (organism == "salm") { 
    df |>      
      select(  
        isolate_id,  
        region,  
        year,   
        month,  
        qtr,   
        sttype,
        final_id_genus, 
        final_id_species,  
        final_id_serotype, 
        cmic_amc,      
        cmic_amp,    
        cmic_azm,    
        cmic_fox,   
        cmic_cro,    
        cmic_chl,    
        cmic_cip,    
        cmic_gen,  
        cmic_nal,  
        cmic_sss, 
        cmic_tet, 
        cmic_sxt, 
        cmic_tio, 
        cmic_str  
        )         
    } else if (organism == "campy") { 
      df |>      
        select(    
          isolate_id, 
          region,    
          year, 
          month, 
          qtr,   
          sttype,
          final_id_genus,
          final_id_species, 
          cmic_gen,    
          cmic_cli,   
          cmic_azm, 
          cmic_ery, 
          cmic_flr,  
          cmic_cip, 
          cmic_nal, 
          cmic_tet,  
          cmic_tel)   
      } else stop("Invalid organism.")    
  }   
# usage of function   
ecoli_amr_data <- select_amr_columns(ecoli_data, "ecoli") salm_amr_data <- select_amr_columns(salm_data, "salm") 
campy_amr_data <- select_amr_columns(ecoli_data, "campy")
```

*Note*: There are likely more eloquent ways to write this function (i.e. `starts_with()` but I initially chose to hard-code the column names. This made it easier to work with a new dataset and explicitly control which variables were included, especially when some columns may no longer be of interest (e.g. `cmic_kan`).

**How it works:**

1.  **Function Definition**
    -   `select_amr_columns` is the name assigned to the function.
2.  **Keyword**
    -   `function()` is telling R that you are creating a new function.
3.  **Parameter**
    -   `df` is the input object the function will perform on.
    -   `organism` is the argument that controls which organism-specific columns should be selected
4.  **Function Body**
    -   `if` and `else_if` are used to check which organism was specified
    -   `if (organism == "ecoli")` select columns specified in `select( )`.
    -   `else_if (organism == "salm")` select columns specified in `select( )`.
    -   `else_if (organism == "campy")` select columns specified in `select( )`.
    -   `else stop ("Invalid organisms")` generates an error message if a user enters an invalid organism (like `"select_amr_columns(ecoli_data, "aeromonas")` ).
5.  **Usage**
    -   `ecoli_amr_data` is the assigned name for a new output (in this case, a dataframe).
    -   `select_amr_columns(ecoli_data, "ecoli")` is applying the function to the value `ecoli_data` , while specifying columns for `"ecoli"`

## Scenario 2 (⭐⭐⭐/5)

Your data may only has MIC variables or you may want to validate the CMIC variables. Rather than repeating most of your code for each MIC variable (there's quite a few antimicrobials on that panel...), you can create a reusable function and apply it across multiple variables - and even across datasets in the future.

``` r
# create function that generates CMIC from MIC  
cmic_expected <- function(df) { 
  mic_cols <- grep("mic_", names(df), value = TRUE)
  for (abx in mic_cols) { 
    x <- str_trim(df[[abx]]) 
    numeric_part <- suppressWarnings(as.numeric( 
      str_extract(x, "[0-9]+\\.?[0-9]*")))  
    df[[paste0("cmic_", sub("mic_", "", abx))]] <- dplyr::case_when(     is.na(x) ~ NA_real_, 
                                                                         str_detect(x, "^\\s*(>|>=|≥)") ~ numeric_part * 2,  
                                                                         str_detect(x, "^\\s*(<|<=|≤)") ~ numeric_part,  
                                                                         !is.na(numeric_part) ~ numeric_part,                                                                         
                                                                         TRUE ~ NA_real_                                                                                                                              )       
    }        
  return(df) 
  }  
# usage of function  
ecoli_amr_data <- cmic_expected(ecoli_data)
```

*Note*: There is a more efficient way to write this function (i.e. `mutate(across())`) but I find reading a loop can be easier and I've personally used parts of this function for data validation and flagging concerns.

**How it works:**

1.  **Function Definition**
    -   `cmic_expected` is the name assigned to the function.
2.  **Keyword**
    -   `function()` is telling R that you are creating a new function.
3.  **Parameter**
    -   `df` is the input object the function will perform on.
4.  **Function Body**
    -   `mic_cols <- grep("mic_", names(df), value = TRUE)` creates a vector containing all column names that start with `"mic_"`. Setting `value = TRUE` keeps the actual column name rather than their positions in the dataframe.
    -   `for (abx in mic_cols) { }` creates a loop that repeats the following **instructions** for item in the `mic_cols` vector. In this case, each item repesents an antimicrobial MIC column and is temporarily called `abx`.
    -   `x <- str_trim(df[[abx]])` extracts the current `abx` column from the dataframe and removes any extra spaces from each observation.
    -   `numeric_part <- suppressWarnings(as.numeric(str_extract(x, "[0-9]+\.?[0-9]*")))` creates a vector called `numeric_part` that extracts only the numeric portion from the observations in `x`.
        -   `suppressWarnings( )` silents the warning messages when values cannot be converted numeric values (e.g. `abc`).
            -   `(as.numeric( ))` converts the extracted values into a numeric variable type.
                -   `str_extract()`extracts the numeric portion from each observation in the column.
    -   `df[[paste0("cmic_", sub("mic_", "", abx))]] <- dplyr::case_when( )` dynamically creates a new column in the dataframe using the {dplyr} package. The new column name is generated by adding `cmic_` to the beginning and removing `"mic_"` from the original column name (e.g. **cmic\_~~mic\_~~amc** ➙ **cmic_amc**). The `case_when ( )` applies a series of conversion rules to the observations:
        -   `is.na(x) ~ NA_real_,` if the orginal value is missing (`na`), return a missing numeric value (`na`).
        -   `str_detect(x, "^\\s*(>|>=|≥)") ~ numeric_part * 2,` if any spaces before text (`^\\s*`) followed by `>` OR `>=` OR `≥` is **detected** in an observation, return the numeric value multiplied by 2.
        -   `str_detect(x, "^\\s*(<|<=|≤)") ~ numeric_part,` if any spaces before text (`^\\s*`) followed by `<` OR `<=` OR `≤` is **detected** in an observation, return the numeric value.
        -   `!is.na(numeric_part) ~ numeric_part,` if observation already contains a valid numeric value with no inequality symbols, return the numeric value unchanged.
        -   `TRUE ~ NA_real_` Any remaining observations that do not match the previous conditions are returned as missing values (`na`).
5.  **Return Statement** - `return(df)` specifies the final output in the function will return.
6.  **Usage**
    -   `ecoli_amr_data` is the assigned name for a new output (in this case, a dataframe).
    -   `cmic_expected(ecoli_data)` is applying the function to the value `ecoli_data`.
