---
title: "Export dataset"
---

```{r}
#| include: false
load("backend/back-datafiles.RData")
```

# xlsx

```{r}
library(writexl)  

write_xlsx(add_rows_data, "data/clean_hockey_data.xlsx") 
```

# csv

```{r}
write.csv(add_rows_data, "data/clean_hockey_data.csv")
```

# R objects

I want to bring attention to saving R objects, as they can be incredibly useful when working on long-term projects. Saving data as an `.xlsx` or `.csv` file can cause you to lose important information from your analysis workflow, such as variable types and other object attributes. Rather than re-running your entire cleaning and wrangling script every time you return to a project, you can save your current progress as an R object and quickly load it back into your environment later.

1.  Save an R object

```{r}
save(add_rows_data, file = "data/clean_hockey_data.Rds")
```

2.  Save multiple R objects

```{r}
save(add_columns_data, add_rows_data, data, file = "data/datafiles.RData")
```

3.  Save the entire workspace

```{r}
save.image(file = "data/work_220526.RData")
```

In all 3 instances, you can `load()` the file by name and continue your coding adventures from where you last left things rather than starting from the beginning.

```{r}
load("data/clean_hockey_data.Rds") 
```
