---
title: "Data Manipulation"
---

```{r}
#| include: false
library(tidyverse)
library(readxl) 

data <- read_excel("data.xlsx")
```

Now, you're probably thinking...

> "This data doesn't follow the best practices mentioned in the first section..."

You're right! Sometimes the data we receive won't be *perfect* - but that's something we can fix!

# Change column cases

**Lower case**

```{r}
lower_case_columns <- rename_all(.tbl=data, .funs=tolower)  

colnames(lower_case_columns)
```

**Upper case**

```{r}
upper_case_columns <- rename_all(.tbl=data, .funs=toupper)  

colnames(upper_case_columns)
```

# Change column names

**One column**

```{r}
one_column_name <- data |>    
  rename("shoting percentage" = SPCT)  

colnames(one_column_name)
```

**Multiple or all columns**

```{r}
all_column_name <- rename(.data = data,
                          "Hockey Player" = Player,
                          "Hockey Team" = Team,
                          "Player Points" = Points,
                          pet = Pet,
                          NHL_division = Division,
                          Shooting_Percentage = SPCT)  

colnames(all_column_name)
```

Notice that when there is a space between words it has to be wrapped with `" "` ? This will become annoying to work with as you continue your data wrangling and analysis.

Again, these names are not ideal BUT this was intentionally done to show you a more sophisticated way to fix column names using the package `janitor`.

**More sophisticated version**

```{r}
#| message: true
#| warning: false

library(janitor)  

clean_names_data <- all_column_name |>   
  clean_names()  

colnames(clean_names_data) 
```

Now, the default is snake_case but if you have a preference - you can modify it.

**For example: camelCase**

```{r}
clean_names_camel_data <- all_column_name |>
  clean_names("lower_camel") 

colnames(clean_names_camel_data)
```

For full transparency, I'll continue these demos with the original column names but change them to lower case.

```{r}
data <- data |>   
  clean_names()  

colnames(data)
```

# Change observations

**Re-code one category within one column**

```{r}
table(data$pet)  
recode_pet_data <- data |> 
  mutate(pet = recode(pet,                      
                      "Snake" = "Lizard"))  

table(recode_pet_data$pet)
```

**Re-code multiple categories within one column**

```{r}
recode_multiple_pet_data <- data |>
  mutate(pet = recode(pet,
                      "Snake" = "Lizard",
                      "Rabbit" = "Horse"))  

table(recode_multiple_pet_data$pet)
```

**Re-code multiple categories within multiple columns**

```{r}
recode_multiple_col_data <- data |>
  mutate(
    pet = recode(pet, "Snake" = "Lizard", "Rabbit" = "Horse"),
    division = recode(division, "PAC" = "PACIF")
    ) 

table(recode_multiple_col_data$division, recode_multiple_col_data$pet)
```

**Re-code multiple columns with similar change**

In this instance, I want to lowercase all my character variables. Basically, `across(where())` is saying... "across the entire dataset where there are character variables, mutate (or change) all observations to lowercase letters.

`toupper` will change all variables to upper case.

```{r}
recode_lowercase_data <- data |>
  mutate(across(where(is.character), tolower))  

head(recode_lowercase_data)
```

Ah! notice that the pet observations didn't change! That's because it's a factor - not a character variable. But, we can change that too! Basically, this is finding the factor variables, changing it to lowercase letters as if it was a character but keeping it as a factor variable.

```{r}
recode_lowercase_fct_data <- data |>
  mutate(across(where(is.factor),
                ~ factor(tolower(as.character(.)))))  

head(recode_lowercase_fct_data)
```

`mutate()` function is incredibly useful and there are a ton more uses that will be discussed later on.

# Select Function

Let's say you are only interested in a few columns, rather than continuing your data processing and analysis with a large dataset, true to its name - simply `select()` the columns you need.

**Select specific columns by name**

```{r}
select_data <- data |>
  select(player, points, pet)

head(select_data)
```

**Select specific columns that all start the same**

```{r}
select_p_data <- data |>
  select(starts_with("p"))  

head(select_p_data)
```

This doesn't have to be a letter, it can be more inclusive or exclusive. Examples include:

`select(starts_with("mic")`

`select(starts_with("cmic_")`

In addition, you can also **select specific columns that all end the same**

`select(ends_with("_AMC")`

**Select specific columns that contain similarities**

```{r}

select_name_contain_data <- data |>
  select(contains("p"))  

head(select_name_contain_data)
```

More relevant example:

`select(contains("mic_"))` this would select columns like **mic_amc** and **cmic_amc**

**Select specific columns by name and similarity**

```{r}
select_name_p_data <- data |> 
  select(team, starts_with("p"))

head(select_name_p_data)
```

Notice that the order of columns changed? This is dependent on the order you select them. If you did `select(starts_with("p"), team)`, you'll notice team is at the end.

Lastly, let's say there's only a few columns you don't want...

**(De-)Select specific columns by name**

```{r}

deselect_name_data <- data |> 
  select(-points, -spct)  

head(deselect_name_data)
```

There are a few ways you can **De-select** columns:

-   For more than two, I'd suggest doing this: `select(-c())`. That way the `-` is coded once.

-   For similar columns, you could apply what you learned above (i.e `select(-starts_with("p"))`

# Filter Function

Let's say you are only interested in specific observations, rather than continuing your data processing and analysis with a large dataset, true to its name - simply `filter()` the observations you need.

**Filter to keep 1 group of interest**

`==` means equal

```{r}

filter_data <- data |>
  filter(pet == "Cat")

head(filter_data)
```

**Filter to remove 1 group**

`!=` means does not equal

```{r}

filter_no_cat_data <- data |>
  filter(pet != "Cat")  

head(filter_no_cat_data)
```

**Filter a numerical range**

`>=` means greater than or equal

`<=` means less than or equal

`>` means greater than

`<` means less than

```{r}
filter_points_data <- data |>
  filter(points >= 15)  

head(filter_points_data)
```

**Filter multiple columns**

Hockey Players that don't own a Cat and have 15 or more points

`&` means and

```{r}
filter_and_data <- data |>
  filter(pet != "Cat" & points >= 15)

head(filter_and_data)
```

Hockey Players that don't own a Cat or have 15 or more points

`|` means or

```{r}
filter_or_data <- data |>
  filter(pet !="Cat" | points >= 15)  

head(filter_or_data)   
```

# Mutate function

The glorious `mutate()` function will become a regular occurrence throughout your coding adventures. Its introduction was teased earlier due to its many capabilities. So far, we've explored how `mutate()` can modify columns, but it can also create entirely new ones!

**Create new binary column**

```{r}
new_bin_col_data <- data |>
  mutate(spct_binary = if_else(spct >= 10, "high", "low"))

head(new_bin_col_data) 
```

`if_else()` is used when there is only two outcomes.

```{r}
new_bin_pet_data <- data |> 
  mutate(pet_binary = if_else(pet == "Cat", 1, 0))

head(new_bin_pet_data)   
```

**Create new categorical column**

```{r}
new_cat_col_data <- data |> 
  mutate(spct_rank = case_when(
    spct < 10 ~ "low",
    spct < 15 ~ "medium",
    TRUE ~ "large"
    )) 

head(new_cat_col_data)
```

`case_when()` is used when there are multiple outcomes.

```{r}
new_cat_pet_data <- data |>   
  mutate(pet_type = case_when(
    pet == "Cat" ~ "indoor",
    pet == "Dog" ~ "outdoor",
    TRUE ~ "other"
    ))  

head(new_cat_pet_data)
```

Now we can add `|` or `&`

**OR**

```{r}

new_bin_or_pet_data <- data |>
  mutate(pet_binary = if_else(pet == "Cat" | pet == "Dog", 1, 0))  

head(new_bin_or_pet_data)
```

If pet is either a cat OR dog - pet_binary = 1, otherwise pet_binary = 0.

```{r}

new_cat_or_pet_data <- data |>
  mutate(pet_type = case_when(
    pet == "Cat" | pet == "Rabbit" ~ "indoor",
    pet == "Dog" ~ "outdoor",
    TRUE ~ "other"
    ))  

head(new_cat_or_pet_data)
```

when pet is either a cat OR rabbit - pet_type = indoor, when pet is a dog - pet_type = outdoor, otherwise pet_type = other.

**AND**

```{r}

new_bin_and_pet_data <- data |>
  mutate(pet_binary = if_else(pet == "Dog" & spct >= 10, 1, 0))  

new_bin_and_pet_data
```

if pet is a dog AND spct is greater than or equal to 10, pet_binary = 1, otherwise pet_binary = 0.

```{r}
new_cat_and_pet_data <- data |>
  mutate(good_vibes = case_when(
    pet == "Dog" & spct >= 10 ~ "decent",
    pet == "Cat" & spct >= 12 ~ "happy",
    TRUE ~ "neutral"
    ))  

new_cat_and_pet_data
```

when pet is a dog AND spct is greater than or equal to 10 - good_vibes = decent,

when pet is a cat AND spct is greater than or equal to 12 - good_vibes = happy,

otherwise, good_vibes = decent.

**Create column based on calculation**

We can make our own like *multiplying*

```{r}
player_spct_double_data <- data |>
  mutate(spct_transform = spct * 2) 

player_spct_double_data
```

or even as fancy as *squaring*

```{r}
player_spct_sq_data <- data |>
  mutate(spct_transform = spct^2)  

player_spct_sq_data
```

Then there's known functions like calculating the *average* by `group_by()` which groups identical observations from one column and calculates the average of another variable within each group.

```{r}
team_spct_data <- data |> 
  group_by(team) |>
  mutate(spct_avg = mean(spct)) 

team_spct_data
```

want standard deviation as well?

```{r}
team_spct_sd_data <- data |>
  group_by(team) |>   
  mutate(spct_avg = mean(spct),
         spct_sd = sd(spct))  

team_spct_sd_data
```

::: {.warning style="padding:0.1em; background-color:#ffb3de; color:#080808"}
<p style="margin-top:1em; text-align:center">

<b>Practice alongside the code - part 2</b>

</p>

<p style="margin-left:1em;">

I didn't lose you yet? phew! 😅 For this next part, if you'd like to continue following along, download the additional dataset below!

<!--Download link -->

<a href="intro-r/data2.xlsx" download="data2.xlsx" class="download-btn">

📥 Download Data

</a>

</p>

<p style="margin-bottom:1em; margin-right:1em; text-align:center; font-family:Georgia">

</p>
:::

# Adding more data to an existing dataframe

## Add columns

1.  Import our new data set!

```{r}
new_columns_data <- read_excel("data2.xlsx", sheet = "Sheet1")  

head(new_columns_data)
```

2.  There are a few `join` functions `prefix_join(x, y)` :

-   `inner_join()` only keeps observations from x that have a matching key in y.

-   `left_join()` keeps all observations from x.

-   `right_join()` keeps all observations from y.

-   `full_join()` keeps all observations from x and y.

```{r}
add_columns_data <- inner_join(data, new_columns_data, 
                               by = c("player" = "Player"))  

add_columns_data 
```

This joins the two datasets together with a shared column **players** and **Players**.

If there were more player observations in **data**, `inner_join()` would join the observations that match, the remainder would not be included.

If there were more player observations in **data**, `left_join()` would join them but those players that are not in `new_columns_data` would have NA observations in **Pregame_drink** and **Country** columns

If there were more player observations in **data**, `right_join()` would join them but would only keep those players that are in `new_column_data`.

If there were more player observations in either **data** or **new_columns_data**, `full_join()` would join them but those players that are not shared in among datasets would have NA observations in **Pregame_drink** and **Country** columns.

Now, let's lowercase these column names before we continue...

```{r}
add_columns_data <- add_columns_data |> 
  clean_names()
```

## Add rows

1.  Import our second new data set!

```{r}
new_rows_data <- read_excel("data2.xlsx", sheet = "Sheet2") |>
  clean_names()  

head(new_rows_data)
```

```{r}
add_rows_data <- bind_rows(add_columns_data, new_rows_data)  

tail(add_rows_data)
```
