---
title: "ggplot"
---

```{r}
#| include: false
library(tidyverse)
```

There are 3 required components:

1.  Data - The data you want to plot.
2.  Mapping (aesthetic) - Tell R what you want to plot.
3.  Geometric representation - Tell R what you want to see in the plot (i.e points, lines, diamonds).

To start, let's load our library and get a dataset.

```{r}
library(tidyverse)  

data()  

Orange
```

You're probably wonder... I thought we were using `ggplot2`. We are... but as mentioned in the introduction to R section, the tidyverse ecosytem is quite resourceful - and include `ggplot2` and for this section, it also has datasets. I chose **Orange**.

Let's run the first two important components...

# Data and Mapping

```{r}
ggplot(data = Orange,
       mapping = aes(x = age, y = circumference))
```

You will notice that the plot appears but there's no values! The `mapping` function sets up your plot *canvas* (x and y axis) but we haven't defined what you want to see in your plot.

# Geometric representation

```{r}

ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference)) +   
  geom_point()
```

Ah! some points are generated 🥳 But let's add some customization...

## Size of points

```{r}
ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference)) +   
  geom_point(size = 3)
```

## Colour of points

```{r}
ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference)) +   
  geom_point(size = 3, color = "orange")
```

## Add a line

(i.e an additional geometric representation)

```{r}
ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference)) +   
  geom_point(size = 3, color = "orange") +   
  geom_line()
```

## Colour of line

```{r}
ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference)) +   
  geom_point(size = 3, color = "orange") +  
  geom_line(color = "darkgreen")
```

We can even specify colour by a categorical variable! By applying the colour in `mapping()`, both `geom_point()` and `geom_line()` will be customized.

```{r}
ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference, color = Tree)) +
  geom_point() +   
  geom_line()
```

Let's say we want to specify colours for **Tree** categories...

We can do this 2 ways:

**1. Hard coding**

```{r}
ggplot(data = Orange,        
       mapping = aes(x = age, y = circumference, color = Tree)) + 
  geom_point() +  
  geom_line() +   
  scale_color_manual(     
    values = c(       
      "1" = "#F5276C",      
      "2" = "darkorange",      
      "3" = "blue",       
      "4" = "purple",       
      "5" = "black" 
      )  
    )
```

**2. List**

```{r}
tree_cols <- c( 
  "1" = "#F5276C", 
  "2" = "darkorange", 
  "3" = "blue", 
  "4" = "purple",
  "5" = "black" 
  )  
ggplot(data = Orange, 
       mapping = aes(x = age, y = circumference, color = Tree)) +
  geom_point() +   
  geom_line() +  
  scale_color_manual(values = tree_cols)
```

Creating a **List** is a far more reproducible. It allows you to apply that same list to other `ggplot()` and the code is visually less daunting.

Now that we know how the important components work, let's simplify the code a bit. The function is smart and knows what to expect in what order so we can remove things!

Let's remove `data`, `mapping`, `x`, and `y` ...

```{r}
ggplot(Orange,         
       aes(age, circumference, color = Tree)) +   
  geom_point() +   
  geom_line() +   
  scale_color_manual(values = tree_cols)
```

# Layers🚧

This component is great for adding additional customization, including (but not limited to) trend lines, axis titles, legends, faceting (multiple plots by category), and other advanced formatting options.
