---
title: "Tables"
---

```{r}
#| include: false
library(tidyverse)

tree_cols <- c( 
  "1" = "#F5276C", 
  "2" = "darkorange", 
  "3" = "blue", 
  "4" = "purple",
  "5" = "black" 
  ) 
```

### 🚧

There are several packages that generate aesthetically appealing tables.

# {flextable}

**Features:**

-   Supports HTML, PDF, Word, and PowerPoint formats.

```{r}
```

# {formatable}

**Features:**

-   Supports HTML formats.

-   Weighted proportions with colour.

-   Display numerical values as percentages.

# {DT}

**Features:**

-   Supports interactive HTML formats.

-   Stripe and hover appearance.

-   Column Filters.

-   Download button.

# {kableExtra}

**Features:**

-   Supports HTML and PDF formats.

-   Stripe and hover appearance

# Tabset

Creating tabs for outputs in HTML.

**Simple**

Let's say you are working on an R markdown report and want to organize your content.

## Orange results

::: panel-tabset
### Plot

```{r}
ggplot(Orange,        
       aes(age, circumference, color = Tree)) +   
  geom_point() +   
  geom_line() +  
  scale_color_manual(values = tree_cols)
```

### Table

```{r}
head(Orange)
```
:::

🚧
